---
title: "HW2 - Gapminder Dataset Exploration"
author: Shuxian Fan
date: Sep 20, 2018
output: github_document
---
## Source Code
```{r ,message=FALSE}
# Install the packages
# install.packages("gapminder")
# install.packages("tidyverse")
# install.packages("ggrepel")
# install.packages("waffle")
# install.packages("ggthemes")
# install.packages("knitr")

# Load the packages and data
library(gapminder)
library(tidyverse)
library(ggplot2)
library(scales)
library(waffle)
library(knitr)
library(cowplot)
```
## Data Description
The `gapminder` object is the main data frame from the package `gapminder`. The data frame has 1704 rows and 6 variables, which is demonstrated as:

```{r}
# return the type of the gapminder object
class(gapminder)

# return the number of columns and rows of the dataframe.
ncol(gapminder)
nrow(gapminder)

# dim for data.frame returns the numbers of rows and columns respectively.
dim(gapminder)
# dimnames(gapminder) will give the names of columns and rows of the dataframe. 
# In this case it will give numbers 1:1704 and the names of the variables.
```
Based on the output, we can answer questions that, gapminder object is a data.frame, also tbl and tbl_df based on its class attribute. 
In addition, there is no missing data in `gapminer` dataset.
```{r}
sum(is.na(gapminder))
```


### Variable Description
The names and data types of the variables are shown as:
```{r}
# get the variable name of the data frame.
names(gapminder)

# extract the class of each variable.
sapply(gapminder, class)
# sapply(gapminder, typeof)
```
The indicated data type can also be shown in the third row of the output of:
```{r}
# get the first parts of the data frame
head(gapminder)
```

Detailed descriptions of the variables can be found in the follwing table:

| Variables  | Description | Type|     
|----------|:-------------------------------------------------|----------|
| `country` | names of the countries, a factor with 142 levels |factor|
|`continent`| names of the continents, a factor with 5 levels|factor|
|`year`| years ranging from 1952 to 2007 with steps of 5 years|integer|
|`lifeExp`| life expectancy in years|numeric|
|`pop`| population| integer|
|`gdpPercap`| GDP per capita (US dollars)| numeric|


There are 142 different countries and 5 continents included in the dataset. The countries and contonents are listed below. 
```{r}
# attach the data so that you can directly use a variable without calling it by $
attach(gapminder)
# Number of levels of the country factor.
length(unique(country))
# This can also be attained by head function.
head(unique(country))

length(unique(continent))
unique(continent)
detach(gapminder)
```
Moreover, the data frame includes the data from 1952 to 2007 with increments 5 years.
```{r}
# Range and of the year variables.
gapminder%>%
  select(year)%>%
range()

unique(gapminder$year)
```
To see how many countries belong to each continent, we can use a table to illustrate as below. We can see that in this dataset, Africa has the maximum number of countries. 
```{r}
attach(gapminder)
# create the table of continent and make it a dataframe
mytable = table(continent)
# display it using kable.
mytable %>%
  as.data.frame()%>%
    mutate(prop = Freq/sum(Freq))%>%
          kable(escape = F, format = "markdown")
detach(gapminder)

```
###  Summary Statistics
From the basic summary statistics of the variables, we can see that the two numeric variables: life expectancy and gdpPercap, range from 23.60 to 82.60 and 241.2 to 113423.1 respectively. Further exploration on the variables and their interactions using various plots are shown in the following section.
```{r}
summary(gapminder)
```

## Exploratory Data Analysis

### Selective Variable Display
In this section, two variables are displayed using a few interesting tools in R packages `ggplot2`,`dplyr`,`waffle`for data vsual display. 

The histograms shown below represent a rough data distribution of the `lifeExp` variable. The first plot shows the count of values of `lifeExp` that lies in some intervals based on the bins we chose. The histogram can show how roughly the data is spread. We can see that the distribution of `lifeExp` is slightly left skewed with most values lie around 70-75. The kernel density plot also clearly shows that there are two humps which indicate that more values lie around these areas. Note that this is just a rough picture of what values can this variable take, we have not used any other information presented in the data yet. This is useful in the sense that when we want to check normality assumption of some variables, histogram can give us some intuition. In our case, `lifeExp` among all continents, countries and years is certainly not normally distributed, which is illustrated by the qq plot below. 
```{r}
theme_set(theme_gray())
# see the common values of the lifeExp variable using a histogram
g1 = ggplot(gapminder, aes(lifeExp))+
  geom_histogram(bins = 30)

# plot the kernel density
g2 = ggplot(gapminder, aes(lifeExp))+
  geom_histogram(bins= 30,aes(y = ..density..))+
  geom_density()

plot_grid(g1,g2, labels = "AUTO",ncol = 2,scale = c(1.,1.))
```



```{r}
# plot the qq plot
ggplot(gapminder, aes(sample = lifeExp))+
  stat_qq()+stat_qq_line()
```

```{r}

```







To illustrate a categorical variable, here I present the barplot, pie chart and a "fresher"" waffle plot of the frequencies of the `continent` variable using the ggplot2 package for data visualization. To start with, we would like to create a data frame based on the table of `continent` we generated earlier as ggplot takes a data frame as an input. As we can see, after make the table as a data frame, now the continent row we had in the table became a column in our new data frame `df`. 

```{r}
# get the table of continent and countries and create a data frame
attach(gapminder)
df = data.frame(table(continent))
head(df)
# add more features to the data frame for further use.
df%>%
mutate(prop = scales::percent(Freq/sum(Freq),2)) -> df

bp = 
  df %>%
  ggplot(aes(x="", y = Freq, fill = continent))+
  geom_bar(width = 1, stat = "identity")+   # add the barplot
  scale_fill_brewer(palette="Set2")         # use brewer color palettes
# Free to try out Dark2, Pastel2, Set1
# A detailed page of different brewer color palettes is attached in the Reference section.
bp
pie = bp + coord_polar("y")
pie
detach(gapminder)
```

```{r}
vals = as.vector(df$Freq)
# create the names of the values
val_names= sprintf("%s (%s)", c("Asia", "Europe", "Africa","Americas", "Oceania "),df$prop)
names(vals) = val_names
waffle::waffle(vals, size = 0.5,rows = 30)
```

### Questions of Interest

Besides seeing how the variables themselves are distributed, our interest also lies on the additional information given by interacting with other variables. For example, what if we want to answer the following questions:

**_Q1: How the population is spread among different continents?_**

Boxplot is a easy way to show how a variable distributed by different groups. For example, we can easily see that among the 5 continents, on average, Asian has the highest population but not as high when it comes to GDP per capita. Oceania has the highest GDP per capita on average. And we can also easily detect outlliers and see how spread out the variables are from the boxplot. 

```{r}
pop_conti = ggplot(gapminder, aes(continent, pop, fill = continent))
pop_conti+
  scale_y_log10()+    # perform the log transformation
  geom_boxplot(outlier.color = "red")+
  scale_fill_brewer(palette="Dark2") # Use the color palette "Dark2"

# There are a few outliers that drag the plot wide that we cannot see the details
# Hereby I remove the observations that have pop >1e10^8 and mainly focus on the major data
gapminder %>%
filter(pop<100000000)%>%
ggplot(aes(pop, fill= continent)) +
  geom_density(alpha=0.4)+
  scale_fill_brewer(palette="Set1")


ggplot(gapminder, aes(continent, gdpPercap,fill = continent))+
  scale_y_log10()+
  geom_boxplot(outlier.color = "red")+
  scale_fill_brewer(palette="Set1")

ggplot(gapminder, aes(log(gdpPercap), fill= continent)) +
  geom_density(alpha=0.4)+
  scale_fill_brewer(palette="Set1")
```

**_Q2: Can we get some intuition on how life expectancy relate to the levels of wealth?_**

Taking the log transformation is commonly used to make the highly skewed distributions less skewed. A comparision of after and before transfromation is shown below. From the less skewed plot we can get the intuition that wealthy areas tend to have higher life expectancy than areas with less wealth.

```{r}
# perform the log transformation of the gdpPercap and create a scatter plot
theme_set(theme_gray())
sc1 = gapminder%>% 
  ggplot(aes(log(gdpPercap), lifeExp, color = continent, shape = continent))+
  geom_point()

sc2 = gapminder%>% 
  ggplot(aes(gdpPercap, lifeExp, color = continent, shape = continent))+
  geom_point()

plot_grid(sc1, sc2, labels = "AUTO")  

  
# Add a fitted curve to the points
gapminder %>% 
  ggplot(aes(gdpPercap, lifeExp)) + 
  scale_x_log10()+
  geom_point() + geom_smooth(color = "darkgreen")

# Faceting
gapminder %>% ggplot(aes(gdpPercap, lifeExp)) + scale_x_log10()+
  geom_point() + 
  geom_smooth() + 
  facet_wrap(~continent, ncol=3)
```





**_Q3: If I want to do some analysis, what about the common assumption of normality?_**

Since overall the `lifeExp` does not comply with normal distribution, what if we explore it by the continent?
By factoring the qq plot with continent, we can see that the Oceania seems to be approximately normally distributed.

```{r}
ggplot(gapminder, aes(sample = lifeExp, colour = factor(continent)))+
  stat_qq()+stat_qq_line()
```

**_Q4: I am interested in the continent Oceania, and howo can I get a closer look into the `gdpPercap` by year?_**

From this plot below we can see the over all trend of the gdpPercap over years along with the individual year distribution of the variable.
```{r}
oce = gapminder%>%
  select(continent, year, gdpPercap)%>%       # select only the columns needed
  filter(continent=="Oceania")

oce %>%             # get the data that in the continent Oceania
  ggplot(aes(year, gdpPercap, group = year))+ 
  scale_y_log10()+
  geom_boxplot()

```

**_Q5: I am interested in the country Thailand and Vienam, I would like to see how `lifeExp` and `gdpPercap` spread out with additional information of the population_**

With adding the size and color scale based on population, we can easily tell the relative size of the population at each points in the regular scatter plot. It gives us more information in a very straight way. 

```{r}
gapminder %>% 
  filter(country %in% c("Thailand", "Vietnam"))%>%
  ggplot(aes(log(gdpPercap), lifeExp, shape = country, color = pop))+
  geom_point(aes(size = pop))+ scale_size_area()+
  scale_color_gradient(low = "#0091ff", high = "#f0650e")
```



## Reference and Source.

`gapminder` Data Source<https://www.gapminder.org/data/>

Brewer Palettes: <http://mkweb.bcgsc.ca/brewer/>

## Extra Exercise
From the out put we can see that the first version which uses `==` actually runs in R without giving errors. But the result shows that it only gives half of the subset we need. The thing R does by running this code is, it compares the country in the odd number of rows with `"Rwanda"` and even number of rows with `"Afghanistan"`, and generates the output. 

The right way to do this is using `%in%` or `|` 
```{r}
tb1 = gapminder%>%
filter(country == c("Rwanda", "Afghanistan"))
dim(tb1)
tb1%>%
  head()%>%
knitr::kable(format = "markdown")


tb2 = filter(gapminder, country %in% c("Rwanda", "Afghanistan"))
dim(tb2)
tb2%>% 
  head()%>%
knitr::kable(format = "markdown")


tb3 = filter(gapminder, country == "Rwanda"| country == "Afghanistan")
dim(tb3)
tb3%>% 
  head()%>%
knitr::kable(format = "markdown")

```

